---
title: "Random sampling of cram files (100 per Schizophrenia PDO)"
author: "Mary T. Yohannes"
date: "4/8/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(dplyr)
library(tools)
```

# 1a) read in table with SCHEMA CRAM paths, sample IDs, and capture (SCZ manifest)
```{r}
ds <- read.csv('~/Desktop/Broad/tj/exomecov/data/2021-03-31_gnomad_scz-exome-manifest.txt', sep = "\t", header=T)
```

# 1b) filter on "primary_disease" column to only "Schizophrenia" patients 
```{r}
SCZ_ds <- ds %>% filter(primary_disease == "Schizophrenia") # 21376 samples 
```

# 1c) from the subsetted dataset of "Schizophrenia" patients, extract the unique PDOs 
```{r}
SCZ_pdo <- distinct(SCZ_ds, pdo) # 41 unique PDOs 
```

# 1d) filter original table to samples with the unique SCZ PDOs 
```{r}
ds_filtered_by_pdo <- ds %>% filter(pdo %in% SCZ_pdo$pdo) # 44250 samples 

# sanity check 
table(ds_filtered_by_pdo$pdo %in% SCZ_pdo$pdo)
```

# 1e) filter to file paths that end with .cram 
```{r}
cram_filtered_by_pdo <- subset(ds_filtered_by_pdo,grepl("^.+(.cram)$",cram)) # 28066 samples with .cram path extension 
distinct(cram_filtered_by_pdo, pdo) # 27 pdos 

# number of files paths with extensions other than .cram - 16184 samples in total 
not_cram_filtered_by_pdo <- subset(ds_filtered_by_pdo,!grepl("^.+(.cram)$",cram)) # 16184 samples in total 
distinct(not_cram_filtered_by_pdo, pdo) # 23 pdos -> 9 pdos are in common between "cram_filtered_by_pdo" and "not_cram_filtered_by_pdo"
```

# 1f) Since I didn't have access to all of the files in the orignal table, prior to running the steps above, I filtered out all sample paths with .cram extension (33182) and separated them into ones I have access to and ones I don't have access to using a python script - took 13 hours and 53 minutes to run on my local computer. So instead of running the python script again on the paths obtained in 1e, I decided to filter the paths from 1e using the results from the python script (this is to see which samples, from SCZ unique pdo samples, I had access to and which ones I didn't)
```{r}
# import outputs obtained from running the python script 
access_paths <- read.table('~/Desktop/Broad/tj/exomecov/data/access_paths.txt') # I have access to 26825 sample paths  
no_access_paths <- read.table('~/Desktop/Broad/tj/exomecov/data/no_access_paths.txt') # 6357 sample paths with no access 

# for all files in the original table with .cram extension 
# for SCZ unique pdo samples - subset of the original dataset 
access_filtered_by_pdo <- cram_filtered_by_pdo %>% filter(cram %in% access_paths$V1) # 21745 samples with access across 18 pdos
no_access_filtered_by_pdo <- cram_filtered_by_pdo %>% filter(cram %in% no_access_paths$V1) # 6321 samples with no access across 9 pdos 

# sanity check - all samples with .cram extension obtained from the unique SCZ pdos were included in the cram_paths.txt file (the access python script was run on this file) 
nrow(access_filtered_by_pdo) + nrow(no_access_filtered_by_pdo) == nrow(cram_filtered_by_pdo)

distinct(no_access_filtered_by_pdo, pdo) 
# 9 pdos with no access 
# PDO-7055 (Google buckets of this PDO do not exist), PDO-10439, PDO-11912, PDO-8917, PDO-14779, PDO-17976, PDO-10544, PDO-9490, PDO-9436
```

# 1g) save the no_access cram paths so you can email it to Christina and ask for access 
```{r}
write.csv(no_access_filtered_by_pdo,"~/Desktop/Broad/tj/exomecov/data/no_access_SCZ_pdo.csv", row.names = FALSE)
```

# 1h) Christina said I should have access to all files now except for the samples with PDO-7055 b/c that workspaces seems to have been created w/o any cram files. So I run the python script (script that checks if I have access or not) on the no_access_paths.txt that I got from running the script the first time - checked the access of 6321 cram files. Then I got two outputs, no_access_access_paths.txt and no_access_no_access_paths.txt. Then, I subsetted "no_access_filtered_by_pdo" to the ones I have access to and added that onto "access_filtered_by_pdo" to get a total of samples (26 pdos). Now, the only PDO I don't have access to is PDO-7055. 
```{r}
# import outputs obtained from running the python script on the no_access_paths.txt 
access_paths2 <- read.table('~/Desktop/Broad/tj/exomecov/data/no_access_access_paths.txt') # I have access to new 6212 sample paths  
no_access_paths2 <- read.table('~/Desktop/Broad/tj/exomecov/data/no_access_no_access_paths.txt') # I still don't have access to 145 sample paths 

# subset the SCZ unique pdo samples to samples with new access 
access_filtered_by_pdo2 <- cram_filtered_by_pdo %>% filter(cram %in% access_paths2$V1) # 6207 samples with access across 8 pdos

# no access from the new run 
no_access_filtered_by_pdo2 <- cram_filtered_by_pdo %>% filter(cram %in% no_access_paths2$V1) # still no access to 114 samples under PDO-7055
# sanity check 
summary(filter(cram_filtered_by_pdo, pdo == 'PDO-7055')$cram == no_access_filtered_by_pdo2$cram)  

# have all the files I have access to in one dataframe 
access_merged <- rbind(access_filtered_by_pdo,access_filtered_by_pdo2) # across all the pdos for SCZ, I have access to a total of 27952 samples 
distinct(access_merged, pdo) # access to 26 pdos for SCZ samples 

# final data: access_merged and no_access_filtered_by_pdo2
# sanity check 
dim(cram_filtered_by_pdo)[1] == dim(access_merged)[1] + dim(no_access_filtered_by_pdo2)[1] #TRUE

# sanity check 
access_merged$cram[1] == access_filtered_by_pdo$cram[1] #TRUE
access_merged$cram[21746] == access_filtered_by_pdo2$cram[1] #TRUE
```

# 2a) randomly select 100 cram file paths you have access to for the mosdepth analysis 
```{r}
# how many pdos are accessible? 
unique(access_merged[c("pdo")]) # OR distinct(access_merged, pdo) # 26 pdos 

# extract 100 samples at random and have them as one file  
set.seed(123) # for reproducibility
samples_per_group <- 100 

# for pdos with <100 samples, include everything 
subsample <- access_merged %>%
  group_by(pdo) %>%
  slice(sample(n(), min(samples_per_group, n())))

group_keys(subsample) # print out the groups/pdos
count(subsample) #OR table(subsample$pdo) # how many samples were sampled per group/pdo? 

# sanity check 
sum(count(subsample)$n) == nrow(subsample) #TRUE
```

# 2b) subset table to columns needed for the mosdepth analysis and for annotation later 
```{r}
# table with cram paths and pdo column  
grouped_subsample_cram <- subsample %>% select(pdo, cram) # extract only the cram file paths and pdo group membership 

# add sample ID as a column for annotation later 
grouped_subsample_cram$sample_ID <- NA 
# extract the sample ID from the CRAM paths and add them onto their respective rows under sample_ID column 
for (i in 1:nrow(grouped_subsample_cram)) {
  n <- strsplit(grouped_subsample_cram$cram[i], .Platform$file.sep)[[1]][7]
  grouped_subsample_cram$sample_ID[i] <- n
}

# for the PDO "EXTERNAL", the above code returns NA value due to the formatting of the path so there are no sample IDs assigned for those samples in the produced table. For now, we will remove it 
grouped_subsample_cram <- grouped_subsample_cram %>% filter(pdo != "EXTERNAL")

# sanity check - 25 pdos without EXTERNAL
distinct(grouped_subsample_cram, pdo)
```

# 2c) write out the table in 2b - pdo, cram paths and sample_ID - tab delimited tsv file without quotes around the string entries 
```{r}
write.table(grouped_subsample_cram,"~/Desktop/Broad/tj/exomecov/data/sampled100_per_pdo/pdo25_crampath_sampleID.tsv",sep="\t",row.names=FALSE, quote=FALSE)
```

# 2d) add the "capture" column and produce a new tsv file for plotting later - this step can be merged with the steps in 2b above with a slight adjustment to the mosdepth python script 
```{r}
id_n_capture <- subsample %>% select(capture, cram) # extract only the cram file paths and capture columns - pdo column is automatically added 

# add sample ID as a column for matching purposes later 
id_n_capture$sample_ID <- NA 

# extract the sample ID from the CRAM paths and add them onto their respective rows under sample_ID column 
for (i in 1:nrow(id_n_capture)) {
  n <- strsplit(id_n_capture$cram[i], .Platform$file.sep)[[1]][7]
  id_n_capture$sample_ID[i] <- n
}

# remove pdo "EXTERNAL"
id_n_capture <- id_n_capture %>% filter(pdo != "EXTERNAL")
```

# 2e) write out the file in 2d - pdo, capture, cram paths, and sample_ID - tab delimited tsv file without quotes around the string entries 
```{r}
write.table(id_n_capture,"~/Desktop/Broad/tj/exomecov/data/sampled100_per_pdo/sampleID_capture.tsv",sep="\t",row.names=FALSE, quote=FALSE)
```




# Junk code - for future reference  
```{r}
# split the whole file into individual data frames (per pdo) - 26 data frames in total 
pdo_subsample <- group_split(subsample)

# extract only the cram paths for each group/data frame/pdo 
for (n in seq(nrow(group_keys(subsample))))
{
  group = group_keys(subsample)$pdo
  assign(group[n], pdo_subsample[[n]] %>% select(cram))
}

# save the paths of samples in each pdo group to individual txt files (one txt file per group - total 26 txt files) - the old mosdepth script was run on these files 
for (i in group_keys(subsample)$pdo)
{
  write.table(get(i),paste0("~/Desktop/Broad/tj/exomecov/data/sampled100_per_pdo/sampled100_", i, ".txt"), row.names=FALSE, col.names=FALSE)
}
```


